---
title: "SparseArray objects"
author:
- name: Hervé Pagès
  affiliation: Fred Hutchinson Cancer Research Center, Seattle, WA
date: "Compiled `r doc_date()`;  Modified 5 April 2023"
package: SparseArray
vignette: |
  %\VignetteIndexEntry{SparseArray objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---


# Introduction

`r Biocpkg("SparseArray")` is an infrastructure package that defines the
SparseArray virtual class whose purpose is to be extended by other S4
classes that wish to represent in-memory multidimensional sparse arrays.

The package currently defines two concrete SparseArray subclasses:
COO\_SparseArray and SVT\_SparseArray. Each subclass uses its own internal
representation for the nonzero multidimensional data, the "COO layout"
and the "SVT layout", respectively.

This vignette focuses on SVT\_SparseArray objects.


# Install and load the package

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("SparseArray")
```

```{r, message=FALSE}
library(SparseArray)
```


# SVT\_SparseArray objects

The SVT\_SparseArray class provides an efficient representation of the
nonzero multidimensional data via a novel layout called the "SVT layout".

Note that SVT\_SparseArray objects mimic the behavior of ordinary matrices
or arrays in R as much as possible. In particular, they suppport most of
the "standard array API" defined in base R.

## Construction

SVT\_SparseArray objects can be constructed in many ways. A common way
is to coerce an ordinary matrix or array to SVT\_SparseArray:
```{r}
m <- matrix(0L, nrow=6, ncol=4)
m[c(1:2, 8, 10, 15:17, 24)] <- (1:8)*10L
svt1 <- as(m, "SVT_SparseArray")
svt1

a <- array(0L, 5:3)
a[c(1:2, 8, 10, 15:17, 20, 24, 40, 56:60)] <- (1:15)*10L
svt2 <- as(a, "SVT_SparseArray")
svt2
```
Alternatively, the ordinary matrix or array can be passed to the
`SVT_SparseArray` constructor function:
```{r}
svt1 <- SVT_SparseArray(m)
svt2 <- SVT_SparseArray(a)
```

Note that coercing an ordinary matrix or array to SparseArray is also
supported and will produce the same results:
```{r}
svt1 <- as(m, "SparseArray")
svt2 <- as(a, "SparseArray")
```
This is because, for most use cases, the SVT\_SparseArray representation
is more efficient than the COO\_SparseArray representation, so the former
is usually preferred over the latter.

For the same reason, the `SparseArray` constructor function will also
give the preference to the SVT\_SparseArray representation:
```{r}
svt1 <- SparseArray(m)
svt2 <- SparseArray(a)
```
This is actually the most convenient way to turn an ordinary matrix or
array into an SVT\_SparseArray object.

Coercion back to ordinary matrix or array is supported:
```{r}
as.array(svt1)  # same as as.matrix(svt1)

as.array(svt2)
```

## Accesors

The standard array accessors are supported:
```{r}
dim(svt2)

length(svt2)

dimnames(svt2) <- list(NULL, letters[1:4], LETTERS[1:3])
svt2
```

Some additional accessors defined in the `r Biocpkg("S4Arrays")` /
`r Biocpkg("SparseArray")` framework:
```{r}
type(svt1)

type(svt1) <- "double"
svt1

is_sparse(svt1)

nzcount(svt1)

sparsity(svt1)
```

## Subsetting and subassignment

```{r}
svt2[5:3, , "C"]
```

Like with ordinary arrays in base R, assigning values of type `"double"` to
an SVT\_SparseArray object of type `"integer"` will automatically change the
type of the object to `"double"`:
```{r}
type(svt2)
svt2[5, 1, 3] <- NaN
type(svt2)
```

See `?SparseArray_subsetting` for more information and additional examples.

## Summarization methods (whole array)

The following summarization methods are provided at the moment:
```{r}
anyNA(svt2)

range(svt2, na.rm=TRUE)

mean(svt2, na.rm=TRUE)
```

See `?SparseArray_summarization` for more information and additional examples.

## Other operations on SVT\_SparseArray objects

More operations will be added in the future e.g. `which()`, `is.na()`,
`is.infinite()`, `is.nan()`, etc..., as well as operations from
the `Compare`, `Logic`, `Math`, `Math2`, and `Complex` groups
(see `?S4groupGeneric` for more information).

## Generate a random SVT\_SparseArray object

Two convenience functions are provided for this:
```{r}
randomSparseArray(c(5, 6, 2), density=0.5)

poissonSparseArray(c(5, 6, 2), density=0.5)
```

See `?randomSparseArray` for more information and additional examples.


# SVT\_SparseMatrix objects

## Transposition

```{r}
t(svt1)
```

## Combining objects by rows or columns

Like ordinary matrices in base R, SVT\_SparseMatrix objects can be
combined by rows or columns, with `rbind()` or `cbind()`:
```{r}
svt3 <- poissonSparseMatrix(6, 2, density=0.5)

cbind(svt1, svt3)
```

See `?SparseArray_combine` for more information and additional examples.

## Matrix multiplication and cross-product

Like ordinary matrices in base R, SVT\_SparseMatrix objects can be
multiplied with the `%*%` operator:
```{r}
m4 <- matrix(0L, nrow=5, ncol=6, dimnames=list(letters[1:5], LETTERS[1:6]))
m4[c(2, 6, 12:17, 22:30)] <- 101:117
svt4 <- SparseArray(m4)

svt4 %*% svt3
```

They also support `crossprod()` and `tcrossprod()`:
```{r}
crossprod(svt3)
```

See `?SparseMatrix_mult` for more information and additional examples.

## matrixStats methods

The `r Biocpkg("SparseArray")` package provides memory-efficient col/row
summarization methods for SVT\_SparseMatrix objects:
```{r}
rowVars(svt4, useNames=TRUE)
```

See `?matrixStats_methods` for more information and additional examples.

## `rowsum()` method

A `rowsum()` method is provided:
```{r}
rowsum(svt4, group=c(1:3, 2:1))
```

See `?rowsum_methods` for more information and additional examples.

## Read/write a sparse matrix from/to a CSV file

Use `writeSparseCSV()` to write a sparse matrix to a CSV file:
```{r}
csv_file <- tempfile()
writeSparseCSV(m4, csv_file)
```

Use `readSparseCSV()` to read the file. This will import the data as
an SVT\_SparseMatrix object:
```{r}
readSparseCSV(csv_file)
```

See `?readSparseCSV` for more information and additional examples.


# Comparison with dgCMatrix objects

The nonzero data of a SVT\_SparseArray object is stored in a _Sparse
Vector Tree_. This internal data representation is referred to as
the "SVT layout". It is similar to the "CSC layout" (compressed, sparse,
column-oriented format) used by CsparseMatrix derivatives from
the `r CRANpkg("Matrix")` package, like dgCMatrix or lgCMatrix objects,
but with the following improvements:

- The "SVT layout" supports sparse arrays of arbitrary dimensions.

- With the "SVT layout", the sparse data can be of any type.
  Whereas CsparseMatrix derivatives only support sparse data of
  type `"double"` or `"logical"` at the moment.

- The "SVT layout" imposes no limit on the number of nonzero elements
  that can be stored. With dgCMatrix/lgCMatrix objects, this number
  must be < 2^31.

- Overall, the "SVT layout" allows more efficient operations on
  SVT\_SparseArray objects.

See `?SVT_SparseArray` for more information.


# Learn more

Please consult the individual man pages in the `r Biocpkg("SparseArray")`
package to learn more about SVT\_SparseArray objects and about the
package. A good starting point is the man page for SparseArray
objects: `?SparseArray`


# Session information

```{r}
sessionInfo()
```

